% !TeX root = ../main.tex
% Add the above to each chapter to make compiling the PDF easier in some editors.

\chapter{Introduction}\label{chapter:introduction}

\section{Section}
TODO: Explain the need, why it emerged etc.



Frame of this work: 

Latency comparison of work sharing vs work stealing. 

What are the variables? 

Morsels 
Mailboxing 
Workstealing normal 
Workstealing deque
Workstealing deque + signaling (similar to Mailboxing)
Granulity of the separation?
Comparison to: TBB, SYCL, and OpenCL and OMP

%(Optional) Automatic threadlocalization of the data --> Maybe better affinity? [tbb::affinity_partitioner]


How can I measure the Latency?
Latency is the time it takes from the last interrupt till the start of the execution. 
-> I can measure this pretty exact in my own implementations 
  BUT this is not doable with TBB or SYCL as I do not have that much control over the execution 
  (OPTIONAL) Extend the tbb/tbb.h or parallel\_for to measure as well? 

What metrics are important here? 

Cache: 
     L1-dcache-load-misses    
     L1-dcache-loads          
     L1-dcache-stores         
     L1-icache-load-misses    
     LLC-loads                
     LLC-load-misses          
     LLC-stores            
     LLC-store-misses

Latency:
  perf sched? 
  custom? 
  heating the caches beforehand

Benchmark results: 
  PBBS? 
  Inf loops? 
  Random events? 
  Power consumption
  scalability
  different type of tasks -> different runtimes?
  Resource contestion (when there are multiple events going on can I elastize the threads, ie change the number of them dynamically)
  NUMA effect of morsels




What is my contribution to the field of low latency scheduling?
work stealing with morsels 
Work stealing with signals + morsels 


11.08.2024: 
Todays TODO: 
  1) Write the split deque suggested by Arora, Blumofe, and Plaxton (SPAA, 1998).
  2) Implement the Job class 
    2.1) must be able to invokable\_v
    2.2) must be able to record the latency + execution time %[#Define XYZ and #if #endif]
    2.3) must support morsels (tbb::blocked\_range<size\_t>)
  

  
